## Evaluating DEG workflows

Because the simulation parameters are known,
`def_DEG()` can be used to determine the true DEG status of each gene.
This information can be used to evaluate how well a DEG workflow ranks true DEGs ahead of non-DEGs.

This section uses simulated count data to evaluate DEG detection performance
with the area under the receiver operating characteristic curve (AUC).
An AUC of 0.5 indicates performance similar to random ranking,
whereas values closer to 1 indicate better ranking of true DEGs.

First, generate a simulated count dataset.

```{r sim-eval-data}
set.seed(1)

x <- sim_gene_counts(n_genes = 20000, p_DEG = c(0.01, 0.04), fc_mean = c(4, 4))
is_DEG <- def_DEG(x, fc = 4)

table(is_DEG)
```

The following MA plot highlights the true DEGs in the simulated dataset.

```{r sim-eval-data-viz, echo=FALSE, include=TRUE}
x_norm <- calc_nf(x)
gene_colors <- ifelse(is_DEG, "#B24745", "#ADADAD")

plot_ma(x_norm, col = gene_colors)
```

Next, define a helper function to calculate the AUC from gene-level p-values.
Smaller p-values are converted to larger scores so that genes
with stronger statistical evidence receive higher ranks.

```{r sim-eval-auc-helper}
library(pROC)

calc_auc <- function(is_deg, p_value) {
    p_value[is.na(p_value)] <- 1
    ranking_score <- -p_value
    roc_obj <- roc(is_deg, ranking_score)
    auc_value <- auc(roc_obj)
    as.numeric(auc_value)
}
```


### Evaluating the edgeR workflow

This subsection compares a standard `r pkg("edgeR", "bioc")` workflow
with a workflow that uses DEGES normalization before `r pkg("edgeR", "bioc")` DEG testing.

First, run TMM normalization and a quasi-likelihood test with `r pkg("edgeR", "bioc")`,
then compute the AUC.

```{r sim-eval-edger-edger}
library(edgeR)

design <- model.matrix(~ group, data = x@exp_design)
d1 <- DGEList(counts = x@data, group = x@exp_design$group)
d1 <- calcNormFactors(d1)
d1 <- estimateDisp(d1, design)
fit1 <- glmQLFit(d1, design)
qlf1 <- glmQLFTest(fit1, coef = 2)
edger_output1 <- topTags(qlf1, n = nrow(d1$counts), sort.by = "none")$table

calc_auc(is_DEG, edger_output1$PValue)
```

Next, estimate normalization factors with the DEGES approach using `calc_nf()`.
The result is converted to a `r pkg("edgeR", "bioc")` object for downstream DEG testing,
and the AUC is computed from the resulting p-values.

```{r sim-eval-edger-deges}
x <- calc_nf(x, orm_func = norm_tmm, test_func = test_edger)

design <- model.matrix(~ group, data = x@exp_design)
d2 <- as_DGEList(x)
d2 <- estimateDisp(d2, design)
fit2 <- glmQLFit(d2, design)
qlf2 <- glmQLFTest(fit2, coef = 2)
edger_output2 <- topTags(qlf2, n = nrow(d2$counts), sort.by = "none")$table

calc_auc(is_DEG, edger_output2$PValue)
```


### Evaluating the DESeq2 workflow

The `r pkg("DESeq2", "bioc")` workflows can be evaluated in the same way.
In `r pkg("seqpatch")`, the built-in `test_deseq2()` function uses an LRT
for DEG screening during DEGES-based normalization.
Here, four workflows are compared:

1. RLE normalization, followed by a `r pkg("DESeq2", "bioc")` Wald test.
2. RLE normalization, followed by a `r pkg("DESeq2", "bioc")` LRT.
3. DEGES using RLE normalization and `r pkg("DESeq2", "bioc")` Wald screening, followed by a `r pkg("DESeq2", "bioc")` Wald test.
4. DEGES using RLE normalization and `r pkg("DESeq2", "bioc")` LRT screening, followed by a `r pkg("DESeq2", "bioc")` LRT.


First, run the `r pkg("DESeq2", "bioc")` Wald test and compute the AUC.

```{r sim-eval-deseq2-Wald}
library(DESeq2)

col_data <- x@exp_design
rownames(col_data) <- colnames(x@data)
design = ~ group

dds1 <- DESeqDataSetFromMatrix(countData = x@data, colData = col_data, design = design)
dds1 <- estimateSizeFactors(dds1)
dds1 <- estimateDispersions(dds1)
dds1 <- nbinomWaldTest(dds1)
deseq2_output_1 = as.data.frame(results(dds1))

calc_auc(is_DEG, deseq2_output_1$pvalue)
```

Next, use the same input data to run a `r pkg("DESeq2", "bioc")` LRT.

```{r sim-eval-deseq2-LRT}
dds2 <- DESeqDataSetFromMatrix(countData = x@data, colData = col_data, design = design)
dds2 <- estimateSizeFactors(dds2)
dds2 <- estimateDispersions(dds2)
dds2 <- nbinomLRT(dds2, full = ~ group, reduced = ~ 1)
deseq2_output_2 = as.data.frame(results(dds2))

calc_auc(is_DEG, deseq2_output_2$pvalue)
```

Next, evaluate a DEGES workflow that uses a Wald test during DEG screening.
Because `r pkg("seqpatch")` does not provide a built-in Wald screening function for `calc_nf()`,
the example first defines `test_deseq2_wald()`.

```{r sim-eval-deseq2-wald-deges}
test_deseq2_wald <- function(x, exp_design, nf = NULL, ...) {
    if (is.null(nf)) nf <- rep(1, ncol(x))

    x <- as.matrix(x)
    exp_design <- as.data.frame(exp_design)
    rownames(exp_design) <- colnames(x)

    effective_lib_sizes <- nf * colSums(x)

    dds <- DESeqDataSetFromMatrix(
        countData = x,
        colData = exp_design,
        design = ~ group
    )
    sizeFactors(dds) <- effective_lib_sizes / mean(effective_lib_sizes)
    dds <- estimateDispersions(dds)
    dds <- nbinomWaldTest(dds)

    test_output <- as.data.frame(results(dds))
    p <- test_output$pvalue
    p[is.na(p)] <- 1
    q <- stats::p.adjust(p, method = "BH")

    data.frame(
        p_value = p,
        q_value = q,
        row.names = NULL,
        check.names = FALSE
    )
}


x3 <- calc_nf(x, norm_func = norm_rle, test_func = test_deseq2_wald)
dds3 <- as_DESeqDataSet(x3)
dds3 <- estimateDispersions(dds3)
dds3 <- nbinomWaldTest(dds3)
deseq2_output_3 = as.data.frame(results(dds3))

calc_auc(is_DEG, deseq2_output_3$pvalue)
```

Finally, evaluate the DEGES workflow that uses the `r pkg("DESeq2", "bioc")`
LRT for DEG screening.

```{r sim-eval-deseq2-lrt-deges}
x4 <- calc_nf(x, norm_func = norm_rle, test_func = test_deseq2)
dds4 <- as_DESeqDataSet(x4)
dds4 <- estimateDispersions(dds4)
dds4 <- nbinomLRT(dds4, full = ~ group, reduced = ~ 1)
deseq2_output_4 = as.data.frame(results(dds4))

calc_auc(is_DEG, deseq2_output_4$pvalue)
```

Finally, summarize the AUC values across the four workflows.


```{r sim-eval-deseq2-output-summary}
data.frame(
    pipeline = c("RLE + Wald", "RLE + LRT", "DEGES(Wald) + Wald", "DEGES(LRT) + LRT"),
    auc = c(
        calc_auc(is_DEG, deseq2_output_1$pvalue),
        calc_auc(is_DEG, deseq2_output_2$pvalue),
        calc_auc(is_DEG, deseq2_output_3$pvalue),
        calc_auc(is_DEG, deseq2_output_4$pvalue)
    )
)
```
