
## Simulating time-series data

The `sim_gene_ts_counts()` function generates RNA-seq count data for time-series experiments.
Each gene is assigned a temporal expression pattern,
and replicate counts are sampled at the specified time points.
As with `sim_gene_counts()`, the result is returned as a `SeqCountData` object.

### Basic time-series simulation

The default call generates 10,000 genes measured at eight time points,
with three biological replicates per time point.
Genes are distributed equally among eight temporal patterns,
including increasing, decreasing, transient, oscillatory, and null patterns.

```{r sim-ts-standard}
set.seed(1)

x_ts <- sim_gene_ts_counts()
x_ts
```

The `exp_design` slot records the time point and biological replicate associated with each sample.
The assigned temporal patterns and their realized proportions are stored in
`x_ts@meta$ts_truth`.

```{r sim-ts-standard-truth}
head(x_ts@exp_design)
table(x_ts@meta$ts_truth$pattern)
x_ts@meta$ts_truth$realized_p_patterns
```

The temporal fold-change matrix is stored in `x_ts@meta$sim_params@fc`.
Rows correspond to genes, and columns correspond to time points.
The pattern templates and gene-specific dynamic range fold changes
can also be inspected through the simulation metadata.

```{r sim-ts-standard-params}
head(x_ts@meta$sim_params@fc)
x_ts@meta$ts_truth$pattern_templates
head(x_ts@meta$ts_truth$dynamic_fc)
```

The simulated expression profiles can be visualized with `plot_gene_profiles()`.
Here, genes are grouped by their true temporal patterns. Normalized counts are averaged
across biological replicates at each time point and scaled within each gene before plotting.
Light lines represent individual genes, and the darker line in each panel represents
the mean profile for that temporal pattern.

```{r sim-ts-standard-profiles}
x_ts <- calc_nf(x_ts)
plot_gene_profiles(x_ts, cl = x_ts@meta$ts_truth$pattern)
```

### Customizing sampling designs and temporal patterns

Use `ts` to define the sampling schedule and `n_replicates`
to set the number of biological replicates at each time point.
The time points may be unevenly spaced.
In the following example, the number of replicates differs among sampling times.

```{r sim-ts-custom-design}
set.seed(1)

x_ts <- sim_gene_ts_counts(
    n_genes = 2000,
    ts = c(0, 1, 3, 7, 14, 21),
    n_replicates = c(3, 3, 3, 4, 4, 4)
)

table(x_ts@exp_design$group)
dim(x_ts@data)

x_ts <- calc_nf(x_ts)
plot_gene_profiles(x_ts, cl = x_ts@meta$ts_truth$pattern)
```

The relative frequencies of the temporal patterns are controlled by `p_patterns`.
A named vector may contain any subset of the available patterns.
Omitted patterns are assigned a probability of zero.
This example generates increasing, decreasing, early peak, and null genes.

```{r sim-ts-custom-patterns}
set.seed(1)

x_ts <- sim_gene_ts_counts(
    n_genes = 2000,
    ts = c(0, 2, 4, 8, 12),
    n_replicates = 3,
    p_patterns = c(
        increase = 0.20,
        decrease = 0.20,
        early_peak = 0.20,
        null = 0.40
    )
)

table(x_ts@meta$ts_truth$pattern)

x_ts <- calc_nf(x_ts)
plot_gene_profiles(x_ts, cl = x_ts@meta$ts_truth$pattern)
```

### Controlling dynamic range fold changes

By default, gene-specific dynamic range fold changes are sampled from a lognormal distribution.
The `fc_mean` argument controls their mean,
and `fc_sd` controls their spread on the natural log fold-change scale.

```{r sim-ts-random-fc}
set.seed(1)

x_ts <- sim_gene_ts_counts(
    fc_mean = 6,
    fc_dist = "lognormal",
    fc_sd = 0.5
)

summary(x_ts@meta$ts_truth$dynamic_fc)
```

For a controlled effect size, set `fc_dist = "fixed"`.
All genes assigned to a non-null pattern then receive the same dynamic range fold change,
whereas null genes retain a fold change of 1.

```{r sim-ts-fixed-fc}
set.seed(1)

x_ts <- sim_gene_ts_counts(
    fc_mean = 4,
    fc_dist = "fixed"
)

table(x_ts@meta$ts_truth$dynamic_fc)
```

Sample-specific library-size variation is controlled by `cv_libsize`.
The realized factors are rescaled to an arithmetic mean of 1 and stored in
`x_ts@meta$libsize_factors`.

```{r sim-ts-library-size}
set.seed(1)

x_ts <- sim_gene_ts_counts(cv_libsize = 0.3)
summary(x_ts@meta$libsize_factors)
```
