## Simulating group comparison data

### Basic group comparison simulation

The simplest call to `sim_gene_counts()` creates
a two-group count matrix containing 10,000 genes with three replicates per group.
The function returns a `SeqCountData` object whose `data` slot contains the count matrix
and whose `exp_design` slot contains the sample grouping.

```{r sim-std-tgc}
library(seqpatch)
set.seed(1)

x <- sim_gene_counts()
x
```

The simulation parameters are stored in `x@meta$sim_params`.
These parameters can be inspected through the object's slots.


```{r sim-std-tgc-simparams}
x@meta$sim_params@n_genes
x@meta$sim_params@p_DEG
head(x@meta$sim_params@params)
head(x@meta$sim_params@fc)
```

Because the true mean and fold-change parameters are known,
differentiall expressed gene (DEG) status can be defined directly from the simulated fold change matrix.
The `def_DEG()` function marks genes as true DEGs when their largest between-group fold change
is at least the cutoff specified by `fc`.
The default cutoff is 2.

```{r sim-std-tgc-degdef}
is_DEG <- def_DEG(x)
head(is_DEG)
```

An MA plot provides a direct view of the simulated counts and the known true DEGs.
Because `plot_ma()` uses normalized counts, estimate normalization factors before plotting.
Here, simulated DEGs are highlighted in red, whereas non-DEGs are shown in gray.

```{r sim-std-tgc-viz}
x <- calc_nf(x)

gene_colors <- ifelse(is_DEG, "#B24745", "#ADADAD")
plot_ma(x, col = gene_colors)
```


The number of genes and the degree of DEG imbalance can be changed through the corresponding arguments.
In the following example, a larger proportion of DEGs is assigned to the second group,
creating an asymmetric expression pattern between the groups.

```{r sim-std-tgc-args}
set.seed(1)

x <- sim_gene_counts(
    n_genes = 2000,
    n_replicates = c(4, 4),
    p_DEG = c(0.05, 0.15),
    fc_mean = c(8, 2),
    group_names = c("control", "stress")
)

is_DEG <- def_DEG(x, fc = 2)

x <- calc_nf(x)

gene_colors <- ifelse(is_DEG, "#B24745", "#ADADAD")
plot_ma(x, col = gene_colors)
```


To simulate a comparison with three groups, provide a replicate count, DEG proportion,
and mean fold change for each group.
The example below creates three groups with different proportions of DEGs.

```{r sim-std-mgc}
set.seed(1)

x <- sim_gene_counts(
    n_replicates = c(3, 3, 3),
    p_DEG = c(0.05, 0.10, 0.15),
    fc_mean = c(4, 4, 4),
    group_names = c("leaf", "root", "shoot")
)
x
```


### Randomizing DEG fold changes

#### Lognormal fold changes

By default, `sim_gene_counts()` samples fold changes for genes assigned as DEGs
from a lognormal distribution.
This produces heterogeneous effect sizes, which more closely resemble typical RNA-seq datasets
than simulations that assign the same fold change to every DEG.

With randomized fold changes, `fc_mean` is interpreted as the mean fold change for genes assigned as DEGs.
For this distribution, `fc_sd` is the standard deviation on the natural log fold-change scale
and controls the spread of the sampled values.


```{r sim-random-fc-lognormal}
set.seed(1)

x <- sim_gene_counts(
    fc_dist = "lognormal",
    fc_mean = c(4, 4),
    fc_sd = 0.5,
    group_names = c("control", "stress")
)

true_fc <- x@meta$sim_params@fc[, "stress"]
summary(true_fc[true_fc != 1])
```

The sampled fold-change matrix is stored in `x@meta$sim_params@fc`.
The `def_DEG()` function uses these fold changes to define the true DEG status.
Because this classification is based on a fold-change cutoff,
the resulting number of true DEGs may differ from the proportion specified by the `p_DEG` argument.


```{r sim-random-fc-viz}
is_DEG <- def_DEG(x, fc = 2)
table(is_DEG)

x <- calc_nf(x, iter = 0)

gene_colors <- ifelse(is_DEG, "#B24745", "#ADADAD")
plot_ma(x, col = gene_colors) + ggtitle("lognormal")
```


#### Gamma fold changes

The gamma distribution provides another option for generating positive fold changes.
Here, the mean fold change is again set by `fc_mean`,
while `fc_sd` controls the coefficient of variation.

```{r sim-random-fc-gamma}
set.seed(1)

x <- sim_gene_counts(
    fc_dist = "gamma",
    fc_mean = c(4, 4),
    fc_sd = 0.5,
    group_names = c("control", "stress")
)

true_fc <- x@meta$sim_params@fc[, "stress"]
summary(true_fc[true_fc != 1])

x <- calc_nf(x, iter = 0)

is_DEG <- def_DEG(x, fc = 2)
gene_colors <- ifelse(is_DEG, "#B24745", "#ADADAD")
plot_ma(x, col = gene_colors) + ggtitle("gamma")
```


#### Fixed fold changes

Fixed fold changes are available for more controlled simulations.
Set `fc_dist = "fixed"` when every DEG assigned to a group
should receive the exact value specified by `fc_mean`.

```{r sim-fixed-fc}
set.seed(1)

x <- sim_gene_counts(
    fc_dist = "fixed",
    fc_mean = c(4, 4),
    group_names = c("control", "stress")
)

true_fc <- x@meta$sim_params@fc[, "stress"]
unique(true_fc[true_fc != 1])

x <- calc_nf(x, iter = 0)

is_DEG <- def_DEG(x, fc = 2)
gene_colors <- ifelse(is_DEG, "#B24745", "#ADADAD")
plot_ma(x, col = gene_colors) + ggtitle("fixed")
```

### Using custom seed matrices

By default, `r pkg("seqpatch")` uses the bundled `arab` dataset,
an *Arabidopsis thaliana* RNA-seq count matrix,
to construct the population of gene-level mean and variance parameters used for simulation.

```{r sim-load-arab}
data(arab, package = "seqpatch")
head(arab)
```

The `sim_gene_counts()` function can also use a user-supplied count matrix
as the seed population.
The example below uses the soybean count matrix stored in
[`docs/data`](https://github.com/bitdessin/seqpatch/tree/main/docs/data) directory
as the seed matrix.

```{r sim-load-seedmatrix}
seed_counts <- read.table("data/soybean.counts.txt.gz", header = TRUE, sep = "\t", row.names = 1)
seed_counts <- as.matrix(seed_counts[, 1:3])
seed_counts <- seed_counts[rowSums(seed_counts) > 0, , drop = FALSE]
head(seed_counts)
```

The seed matrix can then be passed to `sim_gene_counts()` to generate a new count matrix.

```{r sim-custom-data-generate}
set.seed(1)

x <- sim_gene_counts(seed_counts = seed_counts, fc_mean = c(2, 2))
is_DEG <- def_DEG(x, fc = 2)

x <- calc_nf(x)

gene_colors <- ifelse(is_DEG, "#B24745", "#ADADAD")
plot_ma(x, col = gene_colors)
```
