## Advanced usage

```{r advcs-setup, include=FALSE}
source("_config.R")
```

This section presents advanced DEGES-based normalization workflows with `r pkg("seqpatch")`,
including how to define custom normalization and test functions for `calc_nf()`
and how to pass additional parameters to the DEGES workflow.


```{r advcs-load-packages}
library(edgeR)
library(DESeq2)
library(seqpatch)
library(ggplot2)
```


### Custom normalization functions

A DEGES-based normalization workflow requires both a normalization function
and a DEG screening function.
By default, `r pkg("seqpatch")` uses `norm_tmm()` and `test_edger()`,
which are based on the widely used `r pkg("edgeR", "bioc")` package.
The package also provides `norm_rle()` and `test_deseq2()`, which use
`r pkg("DESeq2", "bioc")` methods. These functions can be supplied through
the `norm_func` and `test_func` arguments of `calc_nf()`.
Users may also provide custom functions that follow the required interfaces.

This section demonstrates how to use a custom normalization function with `calc_nf()`.
The function should accept `x` as a count matrix and `exp_design` as an experimental
design data frame, and return a vector containing one normalization factor per sample.
This interface allows users to implement their own normalization methods.

Here, we define an upper-quartile normalization function and pass it to `calc_nf()`.

```{r advcs-uq-func-def}
norm_uq <- function(x, exp_design, ...) {
    x <- as.matrix(x)
    keep <- rowSums(x > 0) > 0
    x <- x[keep, , drop = FALSE]
    uq <- apply(x, 2, quantile, probs = 0.75, names = FALSE)
    nf <- uq / colSums(x)
    nf <- nf / exp(mean(log(nf)))
    nf
}
```

In this example, we customize only the normalization function.
The test function uses the default `r pkg("edgeR", "bioc")` quasi-likelihood test,
so no custom test function is needed.

After defining the function, load the required packages and prepare the included count matrix
as in the basic workflow.
The prepared count file is available from the
[`docs/data`](https://github.com/bitdessin/seqpatch/tree/main/docs/data) directory
of the `r pkg("seqpatch")` GitHub repository.

```{r advcs-uq-framework, message=FALSE, warning=FALSE}
library(edgeR)
library(baySeq)
library(seqpatch)

count_data <- read.table("data/ath.counts.txt.gz",
                         header = TRUE, sep = "\t", row.names = 1)
keep <- rowSums(edgeR::cpm(count_data) > 5) > (ncol(count_data) / 2)
count_data <- as.matrix(count_data[keep, , drop = FALSE])

group <- sapply(strsplit(colnames(count_data), "_"), "[", 1)
exp_design <- data.frame(group = factor(group))

x <- newSeqCountData(count_data, exp_design = exp_design)
```

Then pass the custom `norm_uq()` function to `calc_nf()`
through the `norm_func` argument.


```{r advcs-uq-framework-calcnf, message=FALSE, warning=FALSE}
x <- calc_nf(x, norm_func = norm_uq)
x@meta$nf
```

The example above implements upper-quartile normalization
to demonstrate how a custom normalization function can be integrated into `calc_nf()`.

However, `r pkg("edgeR", "bioc")` can also perform upper-quartile normalization
through `calcNormFactors()` by setting `method = "upperquartile"`.
The workflow can therefore be changed from TMM to upper-quartile normalization as follows.
Additional arguments supplied to `calc_nf()` are forwarded to the normalization function.
In this example, `method = "upperquartile"` is passed to `calcNormFactors()`.

```{r advcs-uq-framework-edgeR, message=FALSE, warning=FALSE}
x <- newSeqCountData(count_data, exp_design = exp_design)
x <- calc_nf(x, method="upperquartile")
x@meta$nf
```


### Custom test functions

TbT [@TbT] is the original DEGES-based normalization workflow.
It estimates normalization factors using TMM and
screens candidate DEGs with `r pkg("baySeq", "bioc")`.
Because `r pkg("seqpatch")` does not provide a built-in `r pkg("baySeq", "bioc")` screening function,
this section shows how to define a custom test function that reproduces the TbT workflow.

At a minimum, the test function should accept `x` as a count matrix,
`exp_design` as an experimental design data frame,
and `nf` as a vector of normalization factors.
It should return a data frame containing `p_value` and `q_value` columns.
The `p_value` column may contain either *p*-values or another compatible statistic used to rank genes.
The function may also accept additional arguments. For example,
`r pkg("baySeq", "bioc")` uses `samplesize` to control bootstrapping
and `cl` to supply cluster information for parallel computation.
The custom function below defines both arguments and provides default values.


```{r advcs-test-bayseq-def, results='hide', message=FALSE, warning=FALSE}
library(baySeq)
library(parallel)

test_bayseq <- function(x, exp_design, nf, samplesize=10000, cl=NULL, ...) {
    d <- new("countData",
            data = x,
            replicates = exp_design$group,
            groups = list(NDE = rep(1, nrow(exp_design)), DE = exp_design$group),
            libsizes = nf * colSums(x))
    d@annotation <- data.frame(gene_id = 1:nrow(x))
    d <- getPriors.NB(d, estimation = "QL", samplesize = samplesize, cl = cl)
    d <- baySeq::getLikelihoods(d, pET = "BIC", cl = cl)
    test_stats <- topCounts(d, group = "DE", number = nrow(x))
    test_stats <- test_stats[order(test_stats$gene_id), ]
    p <- test_stats$FWER.DE
    q <- test_stats$FDR.DE
    p[is.na(p)] <- 1
    q[is.na(q)] <- 1
    data.frame(
        p_value = p,
        q_value = q,
        row.names = NULL, check.names = FALSE)
}
```

For the normalization step, the built-in `norm_tmm()` function can be used directly.
After defining the test function, load the included count matrix and run normalization.
Here, `samplesize = 100` is used only to keep the example concise.
For practical analyses, use a larger value such as 10,000.


```{r advcs-tbt-framework, results='hide', message=FALSE, warning=FALSE}
library(edgeR) # to use 'cpm' function
library(seqpatch)

count_data <- read.table("data/ath.counts.txt.gz",
                         header = TRUE, sep = "\t", row.names = 1)
keep <- rowSums(edgeR::cpm(count_data) > 5) > (ncol(count_data) / 2)
count_data <- as.matrix(count_data[keep, , drop = FALSE])

group <- sapply(strsplit(colnames(count_data), "_"), "[", 1)
exp_design <- data.frame(group = factor(group))

x <- newSeqCountData(count_data, exp_design = exp_design)

if(require("parallel")) {
    cl <- makeCluster(4)
} else {
    cl <- NULL
}

x <- calc_nf(x, norm_func = norm_tmm, test_func = test_bayseq,
             iter = 1, samplesize = 100, cl = cl)

if (!is.null(cl)) {
    stopCluster(cl)
}
```

Next, use the estimated normalization factors for DEG identification.
Although `test_bayseq()` is primarily intended for DEG screening within `calc_nf()`,
it returns gene-level *p*-values and *q*-values and can therefore be reused
for the final DEG analysis.

```{r advcs-tbt-framework-detection, results='hide', message=FALSE, warning=FALSE}
if(require("parallel")) {
    cl <- makeCluster(4)
} else {
    cl <- NULL
}

test_stats <- test_bayseq(x@data, x@exp_design, x@meta$nf, samplesize = 100, cl = cl)

if (!is.null(cl)) {
    stopCluster(cl)
}

test_stats <- data.frame(gene_name = x@gene_names, test_stats)
```

```{r advcs-tbt-output}
head(test_stats)
```

Then, visualize the test results by highlighting genes with `q_value < 0.05`
in the `r pkg("baySeq", "bioc")` results.

```{r advcs-tbt-viz}
is_deg <- test_stats$q_value < 0.05
gene_colors <- ifelse(is_deg, "#B24745", "#ADADAD")
plot_ma(x, col = gene_colors)
```


### GLM for multi-factor comparison

For experiments with multiple experimental factors,
the statistical model should be chosen to match the biological question.
Different models can be supplied to screen candidate DEGs for specific effects during normalization.

Here, we prepare a count matrix for a multi-factor comparison.
The dataset contains RNA-seq counts from control and treated soybean leaves
collected at several time points after water-deficit treatment [@E-GEOD-69469].
The prepared count data are available in the
[`docs/data`](https://github.com/bitdessin/seqpatch/tree/main/docs/data) directory
of the `r pkg("seqpatch")` GitHub repository.


```{r advcs-mfc-soybean-load-data}
count_data <- read.table("data/soybean.counts.txt.gz", header = TRUE, row.names = 1, sep = "\t")
keep <- rowSums(cpm(count_data) > 5) > ncol(count_data) / 2
count_data <- count_data[keep, , drop = FALSE]

exp_design <- data.frame(
    group = factor(sapply(strsplit(colnames(count_data), "_"), "[", 1)),
    timepoint = factor(sapply(strsplit(colnames(count_data), "_"), "[", 2))
)

x <- newSeqCountData(count_data, exp_design = exp_design)
x
```


By default, `calc_nf()` screens candidate DEGs using the available experimental design.
This approach can exclude genes associated with any modeled additive or interaction effect
during DEGES-based normalization.

```{r cs-mfc-soybean-calcnf-default}
x <- calc_nf(x)
x@meta$nf
```

In some analyses, the normalization step should screen candidate DEGs for a specific factor only.
In that case, pass additional arguments to the test function used by `calc_nf()`.

For example, using TMM normalization and a `r pkg("edgeR", "bioc")` quasi-likelihood test,
the workflow below removes candidate DEGs for the treatment effect while adjusting for time point.

```{r cs-mfc-soybean-calcnf-edgeR-treatment-1}
design <- model.matrix(~ group + timepoint, data = exp_design)
head(design)
```

After confirming that the second column of the design matrix represents the treatment effect,
pass `design` and `coef = 2` to `calc_nf()`.

```{r cs-mfc-soybean-calcnf-edgeR-treatment-2}
x <- calc_nf(x, design = design, coef = 2)
x@meta$nf
```

See the `r pkg("edgeR", "bioc")` documentation for more detailed guidance on model matrices,
coefficients, and contrasts.

`r pkg("DESeq2", "bioc")` uses a different testing interface.
For a likelihood ratio test (LRT), specify a full model and a reduced model.
The test evaluates whether terms present only in the full model explain additional variation.
For example, to test the treatment effect after adjusting for time point, use the following workflow.


```{r cs-mfc-soybean-calcnf-DESeq2}
x <- newSeqCountData(count_data, exp_design = exp_design)
x <- calc_nf(x, norm_func = norm_rle, test_func = test_deseq2, full = ~ group + timepoint, reduced = ~ timepoint)
x@meta$nf
```
