## Basic usage

```{r cs-setup, include=FALSE}
source("_config.R")
```

This section introduces practical DEGES-based normalization workflows with `r pkg("seqpatch")`.
The examples follow the main steps in a typical analysis:
load a count matrix, prepare the experimental design, estimate normalization factors,
and perform differentially expressed gene (DEG) analysis.


First, load the packages used in the examples.

```{r cs-load-packages}
library(edgeR)
library(DESeq2)
library(seqpatch)
library(ggplot2)
```

As with `r pkg("edgeR", "bioc")` and `r pkg("DESeq2", "bioc")`,
`r pkg("seqpatch")` requires raw, unnormalized integer counts as input.
Do not use transcripts per million (TPM), fragments per kilobase of transcript per million mapped reads (FPKM),
counts per million (CPM), normalized counts, or log-transformed expression values.



### Two-group comparison

A two-group comparison is one of the most common RNA-seq experimental designs.
Examples include treatment versus control, mutant versus wild type, and one tissue versus another tissue.
This section explains how to perform a two-group comparison
using `r pkg("seqpatch")` with either `r pkg("edgeR", "bioc")` or `r pkg("DESeq2", "bioc")`.

The prepared count data used in this example is available from the
[`docs/data`](https://github.com/bitdessin/seqpatch/tree/main/docs/data) directory
of the `r pkg("seqpatch")` GitHub repository.
The original dataset is available from Expression Atlas under accession
[E-MTAB-4391](https://www.ebi.ac.uk/gxa/experiments/E-MTAB-4391/).
It profiles *Arabidopsis thaliana* roots grown under iron-deficient and control conditions [@E-MTAB-4391].

```{r cs-tgc-ath-download-dataset}
count_data <- read.table("data/ath.counts.txt.gz", header = TRUE, row.names = 1, sep = "\t")
head(count_data)
```

Genes with little or no expression provide limited statistical information and can reduce power in DEG analysis.
Here, we retain genes with more than 5 CPM in more than half of the samples for downstream analysis.

```{r cs-tgc-ath-filter-genes}
keep <- rowSums(cpm(count_data) > 5) > ncol(count_data) / 2
count_data <- count_data[keep, , drop = FALSE]
dim(count_data)
```

Next, prepare the experimental design.
The experimental condition is stored in a data frame named `exp_design`.
Each row of `exp_design` represents one sample and must correspond to a column of `count_data` in the same order.


```{r cs-tgc-ath-exp_design}
group = sapply(strsplit(colnames(count_data), "_"), "[", 1)
exp_design <- data.frame(group = factor(group))
exp_design
```

The count matrix and experimental design are then combined in a `SeqCountData` object.

```{r cs-tgc-ath-expmx}
x <- newSeqCountData(count_data, exp_design = exp_design)
```

The `calc_nf()` function estimates normalization factors using DEGES,
which repeatedly performs four steps:

1. Estimate provisional normalization factors.
2. Identify candidate DEGs.
3. Exclude those candidate genes.
4. Recalculate the normalization factors.

By default, `calc_nf()` uses TMM normalization and a `r pkg("edgeR", "bioc")`
quasi-likelihood test during DEG elimination.

```{r cs-tgc-ath-calc_nf}
x <- calc_nf(x)
```

The estimated normalization factors are stored in the `meta` slot.
They are combined with the library sizes when fitting the downstream statistical model.

```{r cs-tgc-ath-show-nf}
x@meta$nf
```

To identify DEGs using these normalization factors,
use `as_DGEList()` to convert the current object to a `r pkg("edgeR", "bioc")` `DGEList`,
then fit the statistical model with `r pkg("edgeR", "bioc")`.


```{r cs-tgc-ath-test}
d <- as_DGEList(x)

design = model.matrix(~ group, data = exp_design)

d <- estimateDisp(d, design = design)
fit <- glmQLFit(d, design = design)
qlf <- glmQLFTest(fit, coef = 2)
```

After testing, extract the results for all genes.

```{r cs-tgc-ath-results}
test_stats <- topTags(qlf, n = nrow(d$counts), sort.by = "none")$table
head(test_stats)
```

An MA plot displays the log fold change of each gene against its average expression level.
The following example highlights genes with a false discovery rate (FDR) below 0.01 and an absolute log2 fold change greater than 1.

```{r cs-tgc-ath-maplot}
is_deg <- (test_stats$FDR < 0.01) & (abs(test_stats$logFC) > 1)
gene_colors <- ifelse(is_deg, "#B24745", "#ADADAD")
plot_ma(x, col = gene_colors)
```

The significance thresholds used here are examples.
Appropriate thresholds depend on the study design and should be defined before the results are interpreted.


DEGES-based normalization can also use functions from `r pkg("DESeq2", "bioc")`.
For example, relative log expression (RLE) normalization can be combined with a `r pkg("DESeq2", "bioc")`
test for DEG screening during the elimination step.
Because `calc_nf()` uses `r pkg("edgeR", "bioc")` functions by default,
specify the `r pkg("DESeq2", "bioc")` functions through the `norm_func`
and `test_func` arguments:

```{r cs-tgc-ath-DESeq2-norm}
x <- newSeqCountData(count_data, exp_design = exp_design)
x <- calc_nf(x, norm_func = norm_rle, test_func = test_deseq2)
```

Next, use `as_DESeqDataSet()` to convert the object to a `DESeqDataSet`,
then perform downstream DEG testing with `r pkg("DESeq2", "bioc")`.


```{r cs-tgc-ath-DESeq2-test}
dds <- as_DESeqDataSet(x)

dds <- DESeq2::estimateDispersions(dds)
dds <- DESeq2::nbinomLRT(dds, full = ~ group, reduced = ~ 1)
test_stats <- DESeq2::results(dds)
head(test_stats)
```

With `full = ~ group` and `reduced = ~ 1`,
the likelihood ratio test (LRT) compares a full model containing `group`
with a reduced model that excludes `group`.
It therefore tests whether group membership
explains a significant amount of variation in gene expression.

The results can also be visualized with an MA plot.

```{r cs-tgc-ath-DESeq2-maplot}
is_deg <- (test_stats$padj < 0.05) & (abs(test_stats$log2FoldChange) > 1)
gene_colors <- ifelse(is_deg, "#B24745", "#ADADAD")

ma_plot <- plot_ma(x, col = gene_colors)
ma_plot + ggtitle("Shoot apex versus leaf: DESeq2")
```


The examples above use either `r pkg("edgeR", "bioc")` or `r pkg("DESeq2", "bioc")`
for both DEGES normalization and downstream DEG testing.
These components can also be combined differently. For example, DEGES normalization can use
`r pkg("edgeR", "bioc")` functions before downstream testing with `r pkg("DESeq2", "bioc")`,
or the packages can be used in the opposite order.


### Multi-group comparison

RNA-seq experiments often include more than two groups,
for example when comparing expression across treatments or tissues.
Here, we analyze an RNA-seq count matrix from multiple tissues of *Oryza sativa* Japonica Group
to identify genes whose expression varies among tissues.
The prepared count data is available from the
[`docs/data`](https://github.com/bitdessin/seqpatch/tree/main/docs/data) directory
of the `r pkg("seqpatch")` GitHub repository.
The original dataset is available from Expression Atlas under accession
[E-MTAB-2037](https://www.ebi.ac.uk/gxa/experiments/E-MTAB-2037) [@E-MTAB-2037].


```{r cs-mgc-rice-load-data}
count_data <- read.table("data/rice.counts.txt.gz", header = TRUE, row.names = 1, sep = "\t")
keep <- rowSums(cpm(count_data) > 5) > ncol(count_data) / 2
count_data <- count_data[keep, , drop = FALSE]
head(count_data)

group = sapply(strsplit(colnames(count_data), "_"), "[", 1)
exp_design <- data.frame(group = factor(group))
head(exp_design)
```

The count matrix and experimental design are then combined in a `SeqCountData` object.


```{r cs-mgc-rice-expmx}
x <- newSeqCountData(count_data, exp_design = exp_design)
```

By default, `calc_nf()` uses TMM normalization and a `r pkg("edgeR", "bioc")`
quasi-likelihood test during DEG elimination.
In a multi-group design, the default DEGES-based workflow screens candidate DEGs
across all groups using an omnibus-style test.
If normalization should remove candidate DEGs for a specific comparison,
such as leaf versus seed, see the [Advanced Usage](#advanced-usage) chapter.

```{r cs-mgc-rice-calcnf}
x <- calc_nf(x)
```

Convert the normalized object into a `r pkg("edgeR", "bioc")` `DGEList`,
then perform an omnibus DEG test across tissues.

```{r cs-mgc-rice-edgeR}
d <- as_DGEList(x)

design <- model.matrix(~ group, data = exp_design)
d <- estimateDisp(d, design = design)
fit <- glmQLFit(d, design = design)
qlf <- glmQLFTest(fit, coef = 2:ncol(design))

test_stats <- topTags(qlf, n = nrow(d$counts), sort.by = "none")$table
head(test_stats)
```

`coef = 2:ncol(design)` tests whether each gene differs among the tissue groups.
For a specific tissue comparison, set `coef` accordingly or use the `contrast` argument.
Refer to the `r pkg("edgeR", "bioc")` documentation for detailed usage.



The same multi-group comparison can be performed entirely with `r pkg("DESeq2", "bioc")` functions.
First, estimate DEGES normalization factors using RLE normalization and
`r pkg("DESeq2", "bioc")`-based DEG screening by setting `norm_func = norm_rle`
and `test_func = test_deseq2`. Next, convert the normalized object to a `DESeqDataSet`
and use an LRT to test for expression differences among groups.

```{r cs-mgc-rice-DESeq2}
x <- newSeqCountData(count_data, exp_design = exp_design)

x <- calc_nf(x, norm_func = norm_rle, test_func = test_deseq2)

dds <- as_DESeqDataSet(x, design = ~ group)

dds <- estimateDispersions(dds)
dds <- nbinomLRT(dds, full = ~ group, reduced = ~ 1)
test_stats <- results(dds)
head(test_stats)
```

With `full = ~ group` and `reduced = ~ 1`, the LRT compares a full model containing
the group effect with a reduced model that excludes it. This provides an omnibus test
of whether expression differs among any of the tissue groups.


### Multi-factor comparison

Some RNA-seq experiments include multiple experimental factors,
such as treatment, time point, tissue, genotype, and batch.
A multi-factor model can account for these factors simultaneously
and test whether the effect of one factor depends on another.

Common biological questions include:

* Is there a difference between treatment and control?
* Is there a difference between tissues?
* Does the treatment effect depend on time point?
* Does the genotype effect depend on tissue?
* Within a treatment group, does gene expression differ between time points?

The statistical design should be selected according to the biological question.

In this example, we analyze soybean RNA-seq data from control
and treated leaves collected at several time points after water-deficit treatment [@E-GEOD-69469].
The example count data is available from the
[`docs/data`](https://github.com/bitdessin/seqpatch/tree/main/docs/data) directory
of the `r pkg("seqpatch")` GitHub repository.
The original dataset is available from Expression Atlas under accession
[E-GEOD-69469](https://www.ebi.ac.uk/gxa/experiments/E-GEOD-69469/Results).


```{r cs-mfc-soybean-load-data}
count_data <- read.table("data/soybean.counts.txt.gz", header = TRUE, row.names = 1, sep = "\t")
keep <- rowSums(cpm(count_data) > 5) > ncol(count_data) / 2
count_data <- count_data[keep, , drop = FALSE]
head(count_data)

group = sapply(strsplit(colnames(count_data), "_"), "[", 1)
tp = sapply(strsplit(colnames(count_data), "_"), "[", 2)
exp_design <- data.frame(group = factor(group), timepoint = factor(tp))
head(exp_design)
```

Create a `SeqCountData` object from the prepared count matrix and experimental design.

```{r cs-mfc-soybean-expmx}
x <- newSeqCountData(count_data, exp_design = exp_design)
```

During DEGES-based normalization, candidate DEGs are removed from the data
before the normalization factors are recomputed from the remaining genes.
By default, `calc_nf()` screens candidate DEGs using the available experimental design,
so genes associated with any modeled effect can be excluded from normalization factor estimation.

```{r cs-mfc-soybean-calcnf}
x <- calc_nf(x)
```

If normalization should remove candidate DEGs for a specific comparison,
such as control versus treatment after adjusting for time point,
see the [Advanced Usage](#advanced-usage) chapter.

Then convert the normalized object into a `r pkg("edgeR", "bioc")` `DGEList`
and run a `r pkg("edgeR", "bioc")` quasi-likelihood test for the treatment effect while adjusting for time point;
that is, `coef = 2` in this design.

```{r cs-mfc-soybean-edgeR}
d <- as_DGEList(x)

design <- model.matrix(~ group + timepoint, data = exp_design)

d <- estimateDisp(d, design = design)
fit <- glmQLFit(d, design = design)
qlf <- glmQLFTest(fit, coef = 2)

```

Do not assume that a biological effect is always represented by `coef = 2`.
The coefficient number depends on the factor levels and their reference levels,
the order of terms in the model, and whether the model contains interaction terms.
Before running a statistical test, inspect the design matrix column names.
After confirming the meaning of each coefficient, test the coefficient of interest.
In this example, `coef = 2` corresponds to the treatment effect used above.


```{r cs-mfc-soybean-edgeR-coef}
colnames(design)
```

Finally, extract the results.

```{r cs-mfc-soybean-edgeR-results}
test_stats <- topTags(qlf, n = nrow(d$counts), sort.by = "none")$table
head(test_stats)
```


Here, the analysis compares treatment groups
while accounting for differences among time points.
When the model includes interaction terms,
the coefficient for a main effect describes
that effect only at the reference level of the other factor.
It should not be interpreted as the average effect across all tissues,
time points, or developmental stages.
Therefore, the coefficient or contrast must be selected according
to the specific biological comparison of interest.


The same multi-factor comparison can be performed with `r pkg("DESeq2", "bioc")`.


```{r cs-mfc-soybean-DESeq2}
library(DESeq2)

x <- newSeqCountData(count_data, exp_design = exp_design)
x <- calc_nf(x, norm_func = norm_rle, test_func = test_deseq2)

dds <- as_DESeqDataSet(x, design = ~ group + timepoint)

dds <- estimateDispersions(dds)
dds <- nbinomLRT(dds, full = ~ group + timepoint, reduced = ~ timepoint)
test_stats <- results(dds)
head(test_stats)
```

The interpretation of an LRT is determined by the difference between the full and reduced models.
Here, the reduced model removes all terms containing `group`.
The test therefore asks whether treatment-related terms improve model fit.
It does not represent a single pairwise contrast.
