## Quick example

```{r qs-setup, include=FALSE}
source("_config.R")
```

This section presents a standard DEGES-based normalization workflow with `r pkg("seqpatch")`.
The example shows how to load a count data, define the experimental groups,
estimate normalization factors, convert the result to an `r pkg("edgeR", "bioc")` object,
and fit an `r pkg("edgeR", "bioc")` quasi-likelihood model
for differentially expressed gene (DEG) detection.


```{r qs-workflow, results='hide', message=FALSE, warning=FALSE}
library(edgeR)
library(seqpatch)

# gene expression
count_data <- read.table("data/ath.counts.txt.gz",
                         header = TRUE, sep = "\t", row.names = 1)
keep <- rowSums(cpm(count_data) > 5) > (ncol(count_data) / 2)
count_data <- as.matrix(count_data[keep, , drop = FALSE])

# experimental groups
group <- sapply(strsplit(colnames(count_data), "_"), "[", 1)
exp_design <- data.frame(group = group)

# data preparation
x <- newSeqCountData(count_data, exp_design = exp_design)

# calculate normalization factors
x <- calc_nf(x)

# convert to DGEList object for edgeR
d <- as_DGEList(x)

# statistical test
design <- model.matrix(~ group, data = x@exp_design)
d <- estimateDisp(d, design)
fit <- glmQLFit(d, design)
qlf <- glmQLFTest(fit, coef = 2)

test_stats <- topTags(qlf, n = nrow(d$counts), sort.by = "none")$table
```

```{r qs-output}
head(test_stats)
```

The output table contains one row per gene and reports the estimated log2 fold change (`logFC`),
average log2 counts per million (`logCPM`), quasi-likelihood F-statistic (`F`),
*p*-value (`PValue`), and adjusted *p*-value (i.e., *q*-value) (`FDR`).

The DEG results can be visualized with `plot_ma()` by assigning a color to each gene.
In this example, genes passing the selected false discovery rate (FDR)
and fold change thresholds are highlighted.

```{r qs-output-plot}
is_deg <- (test_stats$FDR < 0.05) & (abs(test_stats$logFC) > 1)
gene_colors <- ifelse(is_deg, "#B24745", "#ADADAD")
plot_ma(x, col = gene_colors)
```
