## TCC compatibility

```{r tcc-setup, include=FALSE}
source("_config.R")
```

This chapter compares normalization factors estimated by `r pkg("seqpatch")`
with those obtained from corresponding workflows in `r pkg("TCC", "bioc")`.
It focuses on normalization factor compatibility and does not compare
the final DEG results.

```{r tcc-packages, message=FALSE, warning=FALSE}
library(edgeR)
library(DESeq2)
library(TCC)
library(seqpatch)
```

### Two-group comparison

This section compares `r pkg("seqpatch")` and `r pkg("TCC", "bioc")`
using a two-group design.
The example uses `hypoData`, a small count matrix distributed with `r pkg("TCC", "bioc")`.
The same count matrix and group labels are supplied to both packages.

```{r tcc-tgc-data}
data(hypoData, package = "TCC")

group <- factor(c(1, 1, 1, 2, 2, 2))
exp_design <- data.frame(group = group)
```

#### edgeR screening

Both `r pkg("seqpatch")` and `r pkg("TCC", "bioc")` support a DEGES workflow
that estimates TMM normalization factors and screens candidate DEGs with an
`r pkg("edgeR", "bioc")` quasi-likelihood test.
With matching DEGES settings, the two packages return identical normalization factors
in this example.

```{r tcc-tgc-tcc-tmm}
tcc_tmm <- new("TCC", hypoData, group)
tcc_tmm <- calcNormFactors(
    tcc_tmm,
    norm.method = "tmm",
    test.method = "edger",
    iteration = 3,
    FDR = 0.1,
    floorPDEG = 0.05
)

tcc_tmm$norm.factors
```


```{r tcc-tgc-seqpatch-tmm}
x_tmm <- newSeqCountData(hypoData, exp_design = exp_design)
x_tmm <- calc_nf(
    x_tmm,
    norm_func = norm_tmm,
    test_func = test_edger,
    iter = 3,
    p_cutoff = 0,
    q_cutoff = 0.1,
    min_rm_genes = 0.05
)
x_tmm@meta$nf
```

The normalization factors are identical in this example.

```{r tcc-tgc-edgeR-nf-diff}
tcc_tmm$norm.factors - x_tmm@meta$nf
```

#### DESeq2 screening

Both `r pkg("seqpatch")` and `r pkg("TCC", "bioc")` also support a DESeq2-based DEGES workflow,
but the default screening test is not identical for the two-group case.
`r pkg("seqpatch")` uses relative log expression (RLE) normalization
with a DESeq2 likelihood ratio test (LRT) during DEGES,
whereas the original `r pkg("TCC", "bioc")` DESeq2 workflow uses RLE normalization with a Wald test.
Consequently, the default DESeq2 workflows are expected to produce different normalization factors
in this example.

```{r tcc-tgc-deseq2}
tcc_deseq2 <- new("TCC", hypoData, group)
tcc_deseq2 <- calcNormFactors(
    tcc_deseq2,
    norm.method = "deseq2",
    test.method = "deseq2",
    iteration = 3,
    FDR = 0.1,
    floorPDEG = 0.05
)
tcc_deseq2$norm.factors
```

```{r tcc-tgc-seqpatch-deseq2-lrt}
x_deseq2_lrt <- newSeqCountData(hypoData, exp_design = exp_design)
x_deseq2_lrt <- calc_nf(
    x_deseq2_lrt,
    norm_func = norm_rle,
    test_func = test_deseq2,
    iter = 3,
    p_cutoff = 0,
    q_cutoff = 0.1,
    min_rm_genes = 0.05
)
x_deseq2_lrt@meta$nf
```

To reproduce the normalization factors from the original `r pkg("TCC", "bioc")`
DESeq2 workflow with `r pkg("seqpatch")`,
define a custom DEG screening function based on the DESeq2 Wald test.
The function follows the `calc_nf()` testing interface:
it receives a count matrix, an experimental design data frame, and normalization factors,
and then returns a data frame containing `p_value` and `q_value`.

```{r tcc-tgc-seqpatch-deseq2-wald}
test_deseq2_wald <- function(x, exp_design, nf = NULL, ...) {
    if (is.null(nf)) nf <- rep(1, ncol(x))

    x <- as.matrix(x)
    exp_design <- as.data.frame(exp_design)
    rownames(exp_design) <- colnames(x)

    effective_lib_sizes <- nf * colSums(x)

    dds <- DESeqDataSetFromMatrix(
        countData = x,
        colData = exp_design,
        design = ~ group
    )
    sizeFactors(dds) <- effective_lib_sizes / mean(effective_lib_sizes)
    dds <- estimateDispersions(dds)
    dds <- nbinomWaldTest(dds)

    test_output <- as.data.frame(results(dds))
    p <- test_output$pvalue
    p[is.na(p)] <- 1
    q <- stats::p.adjust(p, method = "BH")

    data.frame(
        p_value = p,
        q_value = q,
        row.names = NULL,
        check.names = FALSE
    )
}

x_deseq2_wald <- newSeqCountData(hypoData, exp_design = exp_design)
x_deseq2_wald <- calc_nf(
    x_deseq2_wald,
    norm_func = norm_rle,
    test_func = test_deseq2_wald,
    iter = 3,
    p_cutoff = 0,
    q_cutoff = 0.1,
    min_rm_genes = 0.05
)
x_deseq2_wald@meta$nf
```

With the custom Wald screening function,
the normalization factors estimated by `r pkg("seqpatch")` match those produced
by the original `r pkg("TCC", "bioc")` DESeq2 workflow in this example.


```{r tcc-tgc-DESeq2-nf-diff}
tcc_deseq2$norm.factors - x_deseq2_wald@meta$nf
```


### Multi-group comparison

This section compares `r pkg("seqpatch")` and `r pkg("TCC", "bioc")`
using a multi-group design.
The example uses `hypoData_mg`, a simulated dataset distributed with `r pkg("TCC", "bioc")`.
It contains three groups with three replicates per group.


```{r tcc-mgt-data}
data(hypoData_mg, package = "TCC")
group <- factor(c(1, 1, 1, 2, 2, 2, 3, 3, 3))
exp_design <- data.frame(group = group)
```


#### edgeR screening

For this multi-group example, both packages use matching TMM and `r pkg("edgeR", "bioc")`
settings during DEGES.
The estimated normalization factors are therefore identical,
as in the two-group TMM and `r pkg("edgeR", "bioc")` comparison.

```{r tcc-mgc-tcc-tmm}
tcc_tmm <- new("TCC", hypoData_mg, group)
tcc_tmm <- calcNormFactors(
    tcc_tmm,
    norm.method = "tmm",
    test.method = "edger",
    iteration = 3,
    FDR = 0.1,
    floorPDEG = 0.05
)

tcc_tmm$norm.factors
```


```{r tcc-mgc-seqpatch-tmm}
x_tmm <- newSeqCountData(hypoData_mg, exp_design = exp_design)
x_tmm <- calc_nf(
    x_tmm,
    norm_func = norm_tmm,
    test_func = test_edger,
    iter = 3,
    p_cutoff = 0,
    q_cutoff = 0.1,
    min_rm_genes = 0.05
)

x_tmm@meta$nf 
```

```{r tcc-mgc-edgeR-nf-diff}
tcc_tmm$norm.factors - x_tmm@meta$nf
```

#### DESeq2 screening

For the multi-group comparison,
the DESeq2-based DEGES workflow in `r pkg("TCC", "bioc")` uses RLE normalization
and an LRT to screen candidate DEGs.
This matches the default `r pkg("seqpatch")` DESeq2 workflow,
so the normalization factors are expected to be identical with matching DEGES settings.

```{r tcc-mgc-deseq2}
tcc_deseq2 <- new("TCC", hypoData_mg, group)
tcc_deseq2 <- calcNormFactors(
    tcc_deseq2,
    norm.method = "deseq2",
    test.method = "deseq2",
    iteration = 3,
    FDR = 0.1,
    floorPDEG = 0.05
)
tcc_deseq2$norm.factors
```

```{r tcc-mgc-seqpatch-deseq2}
x_deseq2 <- newSeqCountData(hypoData_mg, exp_design = exp_design)
x_deseq2 <- calc_nf(
    x_deseq2,
    norm_func = norm_rle,
    test_func = test_deseq2,
    iter = 3,
    p_cutoff = 0,
    q_cutoff = 0.1,
    min_rm_genes = 0.05
)
x_deseq2@meta$nf
```

```{r tcc-mgc-DESeq2-nf-diff}
tcc_deseq2$norm.factors - x_deseq2@meta$nf
```
